\textbf{A framework for effective application of machine learning to microbiome-based classification problems}

### 2020-01-28 submission/manuscript.tex
Pat's edits

Machine learning (ML) modeling of the human microbiome has the potential
to identify microbial biomarkers and aid in the diagnosis of many
diseases such as inflammatory bowel disease, diabetes, and colorectal
cancer. Progress has been made towards developing ML models that predict
health outcomes using bacterial abundances, but inconsistent adoption of
training and evaluation methods call the validity of these models into
question. Furthermore, there appears to be a preference by many
researchers to favor increased model complexity over interpretability.
To overcome these challenges, we trained seven models that used fecal
16S rRNA sequence data to predict the presence of colonic screen
relevant neoplasias (SRNs; n=490 patients, 261 controls and 229 cases).
We developed a reusable open-source pipeline to train, validate, and
interpret ML models. To show the effect of model selection, we assessed
the predictive performance, interpretability, and training time of
L2-regularized logistic regression, L1 and L2-regularized support vector
machines (SVM) with linear and radial basis function kernels, decision
trees, random forest, and gradient boosted trees (XGBoost). The random
forest model performed best at detecting SRNs with an AUROC of 0.695
{[}IQR 0.651-0.739{]} but was slow to train (83.2 h) and not immediately
interpretable. Despite its simplicity, L2-regularized logistic
regression followed random forest in predictive performance with an
AUROC of 0.680 {[}IQR 0.625-0.735{]}, trained faster (12 min), and was
inherently interpretable. Our analysis highlights the importance of
choosing an ML approach based on the goal of the study, as the choice
will inform expectations of performance and interpretability.

@@

As the number of people represented in human microbiome datasets grow,
there is an increasing desire to use microbiome data to diagnose
disease. However, the structure of the human microbiome is remarkably
variable among individuals to the point where it is often difficult to
identify the bacterial populations that are associated with diseases
using traditional statistical models. For example it is not possible to
classify individuals as having healthy colons or colorectal lesions
using Bray-Curtis distances based on the 16S rRNA gene sequences
collected from fecal samples {[}Figure S1{]}. This variation is likely
due to the ability of many bacterial populations to fill the same niche
such that different populations cause the same disease in different
individuals. Furthermore, a growing number of studies have shown that it
is rare for a single bacterial species to be associated with a disease.
Instead, subsets of the microbiome account for differences in health.
Traditional statistical approaches do not adequately account for the
variation in the human microbiome and typically consider the protective
or risk effects of each bacterial population separately (1). Recently,
machine learning (ML) models have grown in popularity among microbiome
researchers as our ability to sample large numbers of individuals has
grown; such models can effectively account for the interpersonal
microbiome variation and the ecology of disease.

ML models can be used to increase our understanding of the variation in
the structure of existing data and in making predictions about new data.
Researchers have used ML models to diagnose and understand the
ecological basis of diseases such as liver cirrhosis, colorectal cancer,
inflammatory bowel diseases, obesity, and type 2 diabetes (2--19). The
task of diagnosing an individual relies on a rigorously validated model.
However, there are common methodological and reporting problems that
arise when applying ML to such data, that need to be addressed for the
field to progress. These problems include a lack of transparency in
which methods are used and how these methods are implemented; evaluating
models without separate held-out test data; unreported variation between
the predictive performance on different folds of cross-validation; and
unreported variation between cross-validation and testing performances.
Though the microbiome field is making progress to avoid some of these
pitfalls including validating their models on independent datasets (8,
19, 20) and introducing accessible and open-source ML tools (21--24),
more work is needed to improve reproducibility further and minimize
overestimating for model performance.

@@

In this study, we provided steps toward standardization of machine
learning methods for microbiome studies which are often poorly
documented and executed. To showcase a rigorous ML pipeline and to shed
light on how ML model selection can affect modeling results, we
performed an empirical analysis comparing the predictive performance,
interpretability and training times of seven modeling approaches with
the same dataset and pipeline. We built three linear models with
different forms of regularization: L2-regularized logistic regression
and L1 and L2-regularized support vector machines (SVM) with a linear
kernel. We also trained four non-linear models: SVM with radial basis
function kernel, a decision tree, random forest, and gradient boosted
trees. We compared their predictive performance, interpretability, and
training time. To demonstrate the performance of these modeling
approaches and our pipeline on a case study, we used a data from a
previously published study that sought to classify individuals as having
healthy colons or colonic lesions based on the 16S rRNA gene sequences
collected from fecal samples (4). This dataset was selected because it
is a relatively large collection of individuals (N=490) connected to a
clinically significant disease where there is ample evidence that the
disease is driven by variation in the microbiome (2, 4, 5, 32). With
this dataset, we developed an ML pipeline that can be used in many
different scenarios for training and evaluating models. This framework
can be easily applied to other host-associated and environmental
microbiome datasets. We also provided an aspirational rubric for
evaluating the rigor of ML practices applied to microbiome data {[}Table
S1{]} to urge the audience to be diligent with with their study design,
model selection, development, evaluation and interpretation.

@@

First, we interpreted the feature importance of the linear models (L1
and L2-regularized SVM with linear kernel and L2-regularized logistic
regression) using the median rank of absolute feature weights for each
OTU {[}Figure 3{]}. We also reviewed the signs of feature weights to
determine whether an OTU was associated with classifying a subject as
being healthy or having an SRN. It was encouraging that many of the
highest-ranked OTUs were shared across these three models (e.g., OTUs
50, 426, 609, 822, 1239). The great benefit of this approach was knowing
the sign and magnitude of each OTU coefficient in the trained model.
This allowed us to immidiately interpret negative and positive
coefficient signs as protective and risk factors, respectively and the
magnitude as the impact of these factors. However, this approach is
limited to linear models or models with prespecified interaction terms.

@@

We also compared the top 20 OTUs selected by permutation importance in
L2-regularized logistic regression {[}Figure S7{]} and the highest
performing tree-based models, random forest and XGBoost {[}Figure 4{]}.
2 and 5 OTUs, respectively, were shared among the models. These results
indicate that we were able to identify important OTUs that are shared
across the highest performing linear and non-linear models when we use
permutation importance as our interpretation method.

We then evaluated the difference in relative abundances of the top 20
OTUs identified in L2-regularized logistic regression and random forest
models between healthy patients and patients with SRNs {[}Figure S8{]}.
There were minimal differences in relative abundances across OTUs
between different diagnoses. This supports our claim that it is not
possible to differentiate disease vs healthy states by focusing on
individual taxa.

@@

We highlighted the importance of model interpretation to gain greater
biological insights into microbiota-associated diseases. In this study,
we showcased two different interpretation methods: ranking each OTU by
(i) their absolute weights in the trained models and (ii) their impact
on the predictive performance based on permutation importance.
Human-associated microbial communities have complex correlation
structures that create collinearity in the datasets. This can hinder our
ability to reliably interpret models because the feature weights of
correlated OTUs are influenced by one another (37). To capture all
important features, once we identify highly ranked OTUs, we should
review their relationships with other OTUs. These relationships will
help us generate new hypotheses about the ecology of the disease and
test them with follow-up experiments. When we used permutation
importance, we partially accounted for collinearity by grouping
correlated OTUs to determine their impact as a group. We grouped OTUs
that had a perfect correlation with each other; however, we could reduce
the correlation threshold to further investigate the relationships among
correlated features. We are choosing to leave the decision to define
correlation and setting up correlations thresholds to the researchers
for their own microbiome questions. However, we would like to urge them
to investigate the correlation structures of the data to avoid
misinterpreting the models. This is likely to be a particular problem
with shotgun metagenomic datasets where collinearity will be more
pronounced due to many genes being correlated with one another because
they come from the same chromosome. True causal mechanisms (e.g., role
of microbiome in colorectal cancer) cannot be explained solely by the
highest performing machine learning model (38). To identify the true
underlying microbial factors of a disease, it is crucial to follow up on
any correlation analyses with further hypothesis testing and
experimentation for biological validation.

@@

Our framework provides a reproducible structure to investigators wanting
to train, evaluate, and interpret their own ML models to generate
hypotheses regarding which OTUs might be biologically relevant. However,
deploying microbiome-based models to make clinical diagnoses or
predictions is a significantly more challenging and distinct undertaking
(42). For example, we currently lack standardized methods to collect
patient samples, generate sequence data, and report clinical data. We
are also challenged by the practical constraints of OTU-based
approaches. The de novo algorithms commonly in use are slow, require
considerable memory, and result in different OTU assignments as new data
are added (43). Finally, we also need independent validation cohorts to
test the performance of a diagnostic model. To realize the potential for
using ML approaches with microbiome data, it is necessary that we direct
our efforts to overcome these challenges.

@@

\textbf{Figure 2. Generalization and classification performance of ML
models using AUROC values of all cross-validation and testing
performances.} The median AUROC for diagnosing individuals with SRN
using bacterial abundances was higher than chance (depicted by a
horizontal line at 0.50) for all the ML models. The predictive
performance of random forest model was higher than other ML models,
though not significantly (p \textgreater{} 0.01). L2-regularized
logistic regression, XGBoost, L2-regularized SVM with linear and radial
basis function kernel performances were not significantly different from
one another. The boxplot shows quartiles at the box ends and the median
as the horizontal line in the box. The whiskers show the farthest points
that were not outliers. Outliers were defined as those data points that
are not within 1.5 times the interquartile ranges.

>>>>

Machine learning (ML) modeling of the human microbiome has the potential
to identify microbial biomarkers and aid in the diagnosis of many
diseases such as inflammatory bowel disease, diabetes, and colorectal
cancer. Progress has been made towards developing ML models that predict
health outcomes using bacterial abundances, but inconsistent adoption of
training and evaluation methods call the validity of these models into
question. Furthermore, there appears to be a preference by many
researchers to favor increased model complexity over interpretability.
To overcome these challenges, we trained seven models that used fecal
16S rRNA sequence data to predict the presence of colonic screen
relevant neoplasias (SRNs; n=490 patients, 261 controls and 229 cases).
We developed a reusable open-source pipeline to train, validate, and
interpret ML models. To show the effect of model selection, we assessed
the predictive performance, interpretability, and training time of
L2-regularized logistic regression, L1 and L2-regularized support vector
machines (SVM) with linear and radial basis function kernels, decision
trees, random forest, and gradient boosted trees (XGBoost). The random
forest model performed best at detecting SRNs with an AUROC of 0.695
{[}IQR 0.651-0.739{]} but was slow to train (83.2 h) and not inherently
interpretable. Despite its simplicity, L2-regularized logistic
regression followed random forest in predictive performance with an
AUROC of 0.680 {[}IQR 0.625-0.735{]}, trained faster (12 min), and was
inherently interpretable. Our analysis highlights the importance of
choosing an ML approach based on the goal of the study, as the choice
will inform expectations of performance and interpretability.

@@

As the number of people represented in human microbiome datasets grow,
there is an increasing desire to use microbiome data to diagnose
diseases. However, the structure of the human microbiome is remarkably
variable among individuals to the point where it is often difficult to
identify the bacterial populations that are associated with diseases
using traditional statistical models. For example it is not possible to
classify individuals as having healthy colons or screen relevant
neoplasia using Bray-Curtis distances based on the 16S rRNA gene
sequences collected from fecal samples {[}Figure S1{]}. This variation
is likely due to the ability of many bacterial populations to fill the
same niche such that different populations cause the same disease in
different individuals. Furthermore, a growing number of studies have
shown that it is rare for a single bacterial species to be associated
with a disease. Instead, subsets of the microbiome account for
differences in health. Traditional statistical approaches do not
adequately account for the variation in the human microbiome and
typically consider the protective or risk effects of each bacterial
population separately (1). Recently, machine learning (ML) models have
grown in popularity among microbiome researchers as our ability to
sample large numbers of individuals has grown; such models can
effectively account for the interpersonal microbiome variation and the
ecology of disease because they consider the relative abundance of each
bacterial population in the context of others rather than in isolation.

ML models can be used to increase our understanding of the variation in
the structure of existing data and in making predictions about new data.
Researchers have used ML models to diagnose and understand the
ecological basis of diseases such as liver cirrhosis, colorectal cancer,
inflammatory bowel diseases, obesity, and type 2 diabetes (2--19). The
task of diagnosing an individual relies on a rigorously validated model.
However, there are common methodological and reporting problems that
arise when applying ML to such data that need to be addressed for the
field to progress. These problems include a lack of transparency in
which methods are used and how these methods are implemented; evaluating
models without separate held-out test data; unreported variation between
the predictive performance on different folds of cross-validation; and
unreported variation between cross-validation and testing performances.
Though the microbiome field is making progress to avoid some of these
pitfalls including validating their models on independent datasets (8,
19, 20) and introducing accessible and open-source ML tools (21--24),
more work is needed to improve reproducibility further and minimize
overestimating for model performance.

@@

In this study, we provided steps toward standardization of machine
learning methods for microbiome studies which are often poorly
documented and executed. To showcase a rigorous ML pipeline and to shed
light on how ML model selection can affect modeling results, we
performed an empirical analysis comparing the predictive performance,
interpretability, data requirements, and training times of seven
modeling approaches with the same dataset and pipeline. We built three
linear models with different forms of regularization: L2-regularized
logistic regression and L1 and L2-regularized support vector machines
(SVM) with a linear kernel. We also trained four non-linear models: SVM
with radial basis function kernel, a decision tree, random forest, and
gradient boosted trees. We compared their predictive performance,
interpretability, and training time. To demonstrate the performance of
these modeling approaches and our pipeline, we present a case study
using data from a previously published study that sought to classify
individuals as having healthy colons or colonic lesions based on the 16S
rRNA gene sequences collected from fecal samples (4). This dataset was
selected because it is a relatively large collection of individuals
(N=490) connected to a clinically significant disease where there is
ample evidence that the disease is driven by variation in the microbiome
(2, 4, 5, 32). With this dataset, we developed an ML pipeline that can
be used in many different scenarios for training and evaluating models.
This framework can be easily applied to other host-associated and
environmental microbiome datasets. We also provided an aspirational
rubric for evaluating the rigor of ML practices applied to microbiome
data {[}Table S1{]} to urge the audience to be diligent in their study
design and model selection, development, evaluation, and interpretation.

@@

First, we interpreted the feature importance of the linear models (L1
and L2-regularized SVM with linear kernel and L2-regularized logistic
regression) using the median rank of absolute feature weights for each
OTU {[}Figure 3{]}. We also reviewed the signs of feature weights to
determine whether an OTU was associated with classifying a subject as
being healthy or having an SRN. It was encouraging that many of the
highest-ranked OTUs were shared across these three models (e.g., OTUs
50, 426, 609, 822, 1239). The benefit of this approach was knowing the
sign and magnitude of each OTU coefficient in the trained model. This
allowed us to immidiately interpret negative and positive coefficient
signs as protective and risk factors, respectively and the magnitude as
the impact of these factors. However, this approach is limited to linear
models or models with prespecified interaction terms.

@@

We also compared the top 20 OTUs selected by permutation importance in
L2-regularized logistic regression {[}Figure S7{]} and the highest
performing tree-based models, random forest and XGBoost {[}Figure 4{]}.
Two and five OTUs, respectively, were shared among the models. These
results indicate that we were able to identify important OTUs that are
shared across the highest performing linear and non-linear models when
we use permutation importance as our interpretation method.

We then evaluated the difference in relative abundances of the top 20
OTUs identified in L2-regularized logistic regression and random forest
models between healthy patients and patients with SRNs {[}Figure S8{]}.
There were minimal differences in the median relative abundances across
OTUs between different diagnoses. This supports our claim that it is not
possible to differentiate disease versus healthy states by focusing on
individual taxa. The ability for ML models to simultaneously consider
the relative abundances of multiple OTUs and their context dependency is
a great advantage over traditional statistical approaches that consider
each OTU in isolation.

@@

We highlighted the importance of model interpretation to gain greater
biological insights into microbiota-associated diseases. In this study,
we showcased two different interpretation methods: ranking each OTU by
(i) their absolute weights in the trained models and (ii) their impact
on the predictive performance based on permutation importance.
Human-associated microbial communities have complex correlation
structures that create collinearity in the datasets. This can hinder our
ability to reliably interpret models because the feature weights of
correlated OTUs are influenced by one another (37). To capture all
important features, once we identify highly ranked OTUs, we should
review their relationships with other OTUs. These relationships will
help us generate new hypotheses about the ecology of the disease and
test them with follow-up experiments. When we used permutation
importance, we partially accounted for collinearity by grouping
correlated OTUs to determine their impact as a group. We grouped OTUs
that had a perfect correlation with each other; however, we could reduce
the correlation threshold to further investigate the relationships among
correlated features. By our approach, we identified 432 OTUs out of
6,920 that had perfect correlations with at least one other OTU. The
decision to establish correlation thresholds is left to researchers to
implement for their own analyses. Regardless of the threshold,
undestanding the correlation structures within the data is critical to
avoid misinterpreting the models. Such structures are likely to be a
particular problem with shotgun metagenomic datasets where collinearity
will be more pronounced due to many genes being correlated with one
another because they come from the same chromosome. Finally, true causal
mechanisms (e.g., role of microbiome in colorectal cancer) cannot be
explained solely by the highest performing machine learning model (38).
To identify the true underlying microbial factors of a disease, it is
crucial to follow up on any correlation analyses with further hypothesis
testing and experimentation for biological validation.

@@

Our framework provides a reproducible structure for investigators
wanting to train, evaluate, and interpret their own ML models to
generate hypotheses regarding which OTUs might be biologically relevant.
However, deploying microbiome-based models to make clinical diagnoses or
predictions is a significantly more challenging and distinct undertaking
(42). For example, we currently lack standardized methods to collect
patient samples, generate sequence data, and report clinical data. We
are also challenged by the practical constraints of OTU-based
approaches. The de novo algorithms commonly in use are slow, require
considerable memory, and result in different OTU assignments as new data
are added (43). Finally, we also need independent validation cohorts to
test the performance of a diagnostic model. To realize the potential for
using ML approaches with microbiome data, it is necessary that we direct
our efforts to overcome these challenges.

@@

\textbf{Figure 2. Generalization and classification performance of ML
models using AUROC values of all cross-validation and testing
performances.} The median AUROC for diagnosing individuals with SRN
using bacterial abundances was higher than chance (depicted by a
horizontal line at 0.50) for all the ML models. The predictive
performance of random forest model was higher than other ML models,
though not significantly (p \textgreater{} 0.05). L2-regularized
logistic regression, XGBoost, L2-regularized SVM with linear and radial
basis function kernel performances were not significantly different from
one another. The boxplot shows quartiles at the box ends and the median
as the horizontal line in the box. The whiskers show the farthest points
that were not outliers. Outliers were defined as those data points that
are not within 1.5 times the interquartile ranges.
